Case based Arabic Morphological Analysis with Compositional Non-deterministic Automata

### 2010-12-10 trunk/tex/rp.tex
future + prelim

%\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{arabtex}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{times}
\usepackage{caption}
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{color}
\usepackage{rotate}
\usepackage{rotating}
\usepackage{multirow}
%\usepackage[colorlinks=false]{color,hyperref}
\usepackage{color,hyperref}
%\usepackage{amsthm}
%\usepackage{booktabs}
\usepackage{natbib}
\bibpunct{[}{]}{,}{n}{}{;}

@@

%IR tools are similar to a Google search 
%engine where documents of interest are selected based on the 
%relevance to a set of keywords of interest to the user.
%NLP targets the automation of understanding human languages.
%This task is the oldest and most difficult task in the artificial 
%intelligence domain.
%The main difficulty behind NLP comes from the ambiguity of 
%natural languages~\cite{Osm08}.
%While NLP techniques cannot deliver their final target yet, they 
%can deliver analysis of sentences into nouns, verbs, and adjectives.
%They can also narrow the meaning of a word or a phrase based on 
%the context to one of its possible meanings.
%Finally they can parse a sentence into a relation between the 
%discovered entities using abstractions such as verb-name 
%abstractions~\cite{Osm08}.

%Information extraction (IE) targets forming data structures and 
%instances of these data structures out of a collection of documents.
%IE tries to fit the output of IR and NLP to templates of 
%interest to the user.
%Data mining can be applied at the end on the output of the 
%IE process to answer user queries about the input 
%documents~\cite{JHa05}.

%\subsection{Arabic text mining in academy }
%In the following we review the major works on Arabic text mining 
%in academy and industry.

@@

{\bf Goals.} Up to our knowledge, CATARM will be the 
first framework that tends to relational Arabic morphology 
and text mining. 
{\bf (1.)} CATARM will build a novel parametrized linguistic 
computational model of Arabic. 
The model will be implemented using efficient structures. 
 {\bf (2.)} It will be amenable to Arabic specific simplifications 
inferred from the correspondence between the fromal Arabic rules
and the relational user queries.
{\bf (3.)} The model will also be amenable to simplifications 
inferred from statistical features extracted from the document 
sets using language independent techniques such as latent semantic 
indexing~\cite{LSI89} and hidden Markov models (HMM).
%CATARM will augment current digital dictionaries to resolve 
%entity semantics and targeted qualificationsl.
{\bf (4.)} CATARM extracts semantic graphs from each text 
document and then uses set arithmetic and graph analysis techniques
to establish relations between these graphs. {\bf (5.)} CATARM will 
solve at least three interesting relational Arabic text case studies 
presented in Section~\ref{s:design:casestudies}.

@@

The framework now comprises a state of the art
case-based morphological analyzer, ATSarf, 
and computational components that allow the user to
specify his queries as one or more FSM
controlling the morphological analyser.
We explain this approach in details in Section~\ref{s:design:fsm}.
ATSarf along with other computational components, 
extract the chains of narrators from a set of narrations
with above 95 percent accuracy and in less than two minutes.

\begin{figure}[tb]
\center{
\resizebox{.75\columnwidth}{!}
{\includegraphics{figs/narrator_chain_output.pdf}
\label{f:narrators} } 
\caption{Directed acyclic graph representing a partial order 
    relation between narrators extracted using ATSarf}
}
\end{figure}

@@

@@

%The Lead PI worked in the area of software arabization since 1996 
%as he worked on developing Arabization modules for Windows NT and 
%an 
%Arabic string manipulation library~\cite{Zar96}. 
%The Lead PI is also involved in working on an open source 
%project that aims at comprehensive automation of authenticity 
%checks against a corpus of literature heritage~\cite{Zaraket06}.
%The Lead PI is leading an effort at AUB to introduce 
%Computational Arabic in the curriculum of the ECE program. 
%He is also leading a team of students in an effort to place 
%the infrastructure needed for this project to start. 
%The lead PI worked and published about relational logic and 
%logic solvers applied to structured languages as follows.
%\begin{itemize}
%\item Sequential circuits for relational analysis. F. Zaraket, A. Aziz, and S. Khurshid. International Conference on Software Engineering, Minneapolis MN, 2007. 
%\item Sequential circuits for program analysis. F. Zaraket, A. Aziz, and S. Khurshid. Automated Software Engineering, Atlanta GA, 2007. 
%\end{itemize}

@@

@@

%\begin{figure}[tb]
\begin{figure}
\center{
\resizebox{.75\columnwidth}{!}
{\includegraphics{figs/literature_case-study1.pdf}
\label{f:literature}
} 
\caption{CATARM diagram for the literature case study }
}
\end{figure}

@@

The computational components can be direct CATARM components
or custom user components built from other CATARM components.
We feed the hadith books to a user Narrator Chain Extractor
component that we will discuss later in 
Section~\ref{s:design:chainext}. We  obtain a set of chains
of narrators that we feed to a morphological analyzer (MA) 
component controlled by a merger controller. 
The merger controller will look at the morphological analysis
of the narrator names and will decide to merge two narrators in 
one graph node in case they were morphologicaly equal. 
It will also give feedback to the MA component to stop the
analysis when a it reaches a satisfactory result.
We consider this graph as a partial order relation between 
narrators where a narrator precedes another narrator if 
he narrated from him.

@@

%\begin{figure}[tb]
\begin{figure}
\center{
\resizebox{.8\columnwidth}{!}
{\includegraphics{figs/security_case-study.pdf}
\label{f:security}
}
\caption{CATARM diagram for the security case study}
}
\end{figure}

@@

\subsection{CATRAM components}
\label{s:design:components}

@@

\subsection{Methodology}

\begin{figure}
\center{
\resizebox{.4\columnwidth}{!}
{\input{figs/hadith.pdftex_t}
\label{f:statemachine}
}
\caption{State-machine used to detect the chain of narrators}
}
\end{figure}

{\bf Case-based morphological analyzer.}
We experimented with several morphological analyzers.
We think that using finite state machines for morphological analysis
can be implemented in a more efficient way compared to other 
approaches. 
For example the concatenative rules used in Buckwalter~\cite{Tim04}
can be encoded in three FSM with linear execution time using 
the trie~\cite{datries} data structure.

@@

\begin{figure}
\center{
\resizebox{.8\columnwidth}{!}
{\includegraphics{figs/detailed_submachine.pdf}
\label{f:chain}
}
\caption{CATARM detailed diagram for the chain of narrators extractor}
}
\end{figure}

@@

The solver may make use of existing propositional~\cite{MiniSAT} or
higher order logic~\cite{Z3} solvers if needed. 
These solvers can solve satisfiability questions.
We will explore the feasibility and utility 
of extracting a necessary set of 
constraints from the query and answering that with an off-the-shelf
logic solver.

@@

{\bf Development methodology.}
We will integrate all the mentioned techniques in a scalable 
framework for Arabic text analysis.
The framework as a whole will allow for exhausting Arabic text 
analysis and allow for extensions of the framework.
A user should be able to iteratively and repetitively apply the 
abode techniques on his input text.

We will follow an eXtreme Programming 
methodology in the process of building CATARM.
We will use open source tools, libraries, and frameworks to 
build our libraries.
We will build on existing open source programming and text 
mining tools as well as open source Arabic text analysis tools 
based on NLP such as MADA+TOKAN~\cite{Rot08}, 
NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.

@@

The LPI offers a course on computational Arabic attended
by excellent senior and graduate students at AUB. 
Once funded, the team can be directly formed and work can start
with productive team members.

\subsection{Work plan}

@@

We anticipate that the research will lead to the production of 
CATARM, an Arabic text analysis framework that allows relational 
queries between Arabic text documents.
The components of the framework are listed and discussed
in details in Section~\ref{s:design}. 
In the following we will discuss what to anticipate and 
how to evaluate the different components of CATARM.

@@

\subsection{CATRAM evaluation }

@@

Enhance morphological analysis
Look for other optimizations
Generalize the framework to other Arabic-like languages
Generalize the case-based approach 
Integrate other text mining techniques such as vocalization,
suggestions...

>>>>

%\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{arabtex}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{times}
\usepackage{caption}
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{color}
\usepackage{rotate}
\usepackage{rotating}
\usepackage{multirow}
%\usepackage[colorlinks=false]{color,hyperref}
\usepackage{color,hyperref}
%\usepackage{amsthm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{natbib}
\bibpunct{[}{]}{,}{n}{}{;}

@@

@@

{\bf Goals.} Up to our knowledge, CATARM will be the 
first framework that tends to relational Arabic morphology 
and text mining. 
{\bf (1.)} CATARM will build a novel parametrized linguistic 
computational model of Arabic. 
The model will be implemented using efficient structures. 
 {\bf (2.)} It will be amenable to Arabic specific simplifications 
inferred from the correspondence between the fromal Arabic rules
and the relational user queries.
{\bf (3.)} The model will also be amenable to simplifications 
inferred from statistical features extracted from the document 
sets using language independent techniques such as latent semantic 
indexing~\cite{LSI89} and hidden Markov models (HMM).
%CATARM will augment current digital dictionaries to resolve 
%entity semantics and targeted qualificationsl.
{\bf (4.)} CATARM extracts semantic graphs from each text 
document and then uses set arithmetic and graph analysis techniques
to establish relations between these graphs. {\bf (5.)} CATARM will 
solve at least three interesting relational Arabic text case studies 
presented in Sections~\ref{s:design:sec} and~\ref{s:design:lit}.

@@

The framework now comprises a state of the art
case-based morphological analyzer, ATSarf, 
and computational components that allow the user to
specify his queries as one or more FSM
controlling the morphological analyser.
We explain this approach in details in 
Section~\ref{s:design:method}.
ATSarf along with other computational components, 
extract the chains of narrators from a set of narrations
with above 95 percent accuracy and in less than two minutes.

\begin{figure}[tb]
\center{
\resizebox{.75\columnwidth}{!}
{\includegraphics{figs/narrator_chain_output.pdf}}
\caption{Directed acyclic graph representing a partial order 
    relation between narrators extracted using ATSarf}
\label{f:narrators} 
}
\end{figure}

@@

\subsection{PI related work}

The Co-LPI, Rehab Duwairi, has several years of experience 
working on data mining for Arabic documents. 
She initially targeted Arabic text categorization where a 
dataset of 1000 documents where used to train and test 
classifiers~\cite{Duw06,Duw07}.
Word morphology comes to the surface when extracting the 
feature vectors of documents. 
In that work, words were stemmed before using them as features. 
Also, word frequency and inverted term frequency were used to 
quantify the importance of features.

In a recent work, Duwairi, Refae and Khasawneh~\cite{Duw09}
have addressed text categorization again. 
This research investigated possible word forms that can be 
used to create feature vectors of documents in particular: 
original word, stem of the word, light-stem (shallow stemming) 
of the word, and word augmented with a list of synonyms. 
Therefore, every document is represented using four different 
feature vectors, the result revealed that light-stemming is the 
best representation as it preserves the meaning of the words 
and yet reduces the dimensions of feature vectors. 
The dataset size was 15,000 documents that belong to three 
categories.

In recent research efforts, the CO-LPI is investigating 
keyphrase extraction fro Arabic documents.

@@

The Lead PI worked in the area of software arabization since 1996 
as he worked on developing Arabization modules for Windows NT 
and an Arabic string manipulation library~\cite{Zar96}. 
The Lead PI is also involved in working on an open source 
project that aims at comprehensive automation of authenticity 
checks against a corpus of literature heritage~\cite{Zaraket06}.
The Lead PI is leading an effort at AUB to introduce 
Computational Arabic in the curriculum of the ECE program. 
He is also leading a team of students in an effort to place 
the infrastructure needed for this project to start.

The lead PI holds several approved patents and has several 
patents pending approval in the area of logic verification 
and synthesis.
The lead PI worked and published about relational logic and 
logic solvers applied to structured languages as follows.
\begin{itemize}
\item Sequential circuits for relational analysis. F. Zaraket, A. Aziz, and S. Khurshid. International Conference on Software Engineering, Minneapolis MN, 2007. 
\item Sequential circuits for program analysis. F. Zaraket, A. Aziz, and S. Khurshid. Automated Software Engineering, Atlanta GA, 2007. 
\end{itemize}

@@

IR tools are similar to a Google search 
engine where documents of interest are selected based on the 
relevance to a set of keywords of interest to the user.
NLP targets the automation of understanding human languages.
This task is the oldest and most difficult task in the artificial 
intelligence domain.
The main difficulty behind NLP comes from the ambiguity of 
natural languages~\cite{Osm08}.
While NLP techniques cannot deliver their final target yet, they 
can deliver analysis of sentences into nouns, verbs, 
and adjectives.
They can also narrow the meaning of a word or a phrase based on 
the context to one of its possible meanings.
Finally they can parse a sentence into a relation between the 
discovered entities using abstractions such as verb-name 
abstractions~\cite{Osm08}.

IE targets forming data structures and instances of these 
data structures out of a collection of documents.
IE tries to fit the output of IR and NLP to templates of 
interest to the user.
Data mining can be applied at the end on the output of the 
IE process to answer user queries about the input 
documents~\cite{JHa05}.

@@

%\begin{figure}[tb]
\begin{figure}
\center{
\resizebox{.75\columnwidth}{!}
{\includegraphics{figs/literature_case-study1.pdf}
} 
\caption{CATARM diagram for the literature case study }
\label{f:literature}
}
\end{figure}

@@

The computational components can be direct CATARM components
or custom user components built from other CATARM components.
We feed the hadith books to a user Narrator Chain Extractor
component that we will discuss later in 
Section~\ref{s:design:method}. We  obtain a set of chains
of narrators that we feed to a morphological analyzer (MA) 
component controlled by a merger controller. 
The merger controller will look at the morphological analysis
of the narrator names and will decide to merge two narrators in 
one graph node in case they were morphologicaly equal. 
It will also give feedback to the MA component to stop the
analysis when a it reaches a satisfactory result.
We consider this graph as a partial order relation between 
narrators where a narrator precedes another narrator if 
he narrated from him.

@@

%\begin{figure}[tb]
\begin{figure}
\center{
\resizebox{.8\columnwidth}{!}
{\includegraphics{figs/security_case-study.pdf}
}
\caption{CATARM diagram for the security case study}
\label{f:security}
}
\end{figure}

@@

\subsection{CATARM components}
\label{s:design:components}

@@

\subsection{Methodology}
\label{s:design:method}

\begin{figure}
\center{
\resizebox{.4\columnwidth}{!}
{\input{figs/hadith.pdftex_t}
}
\caption{State-machine used to detect the chain of narrators}
\label{f:statemachine}
}
\end{figure}

{\bf Case-based morphological analyzer.}
We experimented with several morphological analyzers.
We think that using finite state machines for morphological analysis
can be implemented in a more efficient way compared to other 
approaches. 
For example the concatenative rules used in Buckwalter~\cite{Tim04}
can be encoded in three FSM with linear execution time using 
the trie~\cite{Aoe:89} data structure.

@@

\begin{figure}
\center{
\resizebox{.8\columnwidth}{!}
{\includegraphics{figs/detailed_submachine.pdf}
}
\caption{CATARM detailed diagram for the chain of narrators extractor}
\label{f:chain}
}
\end{figure}

@@

The solver may make use of existing 
propositional~\cite{MiniSAT04} or
higher order logic~\cite{Z308} solvers if needed. 
These solvers can solve satisfiability questions.
We will explore the feasibility and utility 
of extracting a necessary set of 
constraints from the query and answering that with an off-the-shelf
logic solver.

@@

@@

The LPI offers a course on computational Arabic attended
by excellent senior and graduate students at AUB. 
Once funded, the team can be directly formed and work can start
with readily productive team members.

\subsection{Work plan}
\label{s:design:plan}

%\begin{table}[bt]
\begin{table}[bt]
\centering
\caption{Three year quarter based work plan.}
%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
%\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
\small
\begin{tabular}{lp{.1cm}ccccp{.1cm}ccccp{.1cm}cccc} \\
& & \multicolumn{4}{c}{Year 1} & & \multicolumn{4}{c}{Year 2} & & \multicolumn{4}{c}{Year 3} \\ \cline{3-6} \cline{8-11} \cline{13-16} %\\
    %\cline{2-13} \\ 
& & Q1 & Q2 & Q3 & Q4 & & Q1 & Q2 & Q3 & Q4 & & Q1 & Q2 & Q3 & Q4 \\ \bottomrule

Case based & & 
W & W & W & E & & E & E & E & E & & E & E & E & E \\ 
morphological analyzer & & 
 &  & D & DM & &   & M & M & M & & M & M & M & M \\  \hline

\multirow{2}{*}{Arabic LCM} & & 
P & W & W & W & & W & E & E & E & & E & E & E & E \\ 
& & 
 &  &  &  & & D & DM & M & M & & M & M & M & M \\ \hline

\multirow{2}{*}{Query language} & & 
P & P & W & W & & W & W & E & E & & E & E & E & E \\ 
& & 
  &  &  &  & &  & D & DM & M & & M & M & M & M \\ \hline

Language & & 
P & P & P & W & & W & W & W & W & & E & W & W & W \\ 
optimizations & & 
 &  &  &  & &  &  &  & D & & DM & M & M & D \\ \hline

Data  & & 
P & P & P & W & & W & W & W & W & & E & W & W & W \\ 
optimizations & & 
 &  &  &  & &  &  &  & D & & DM & M & M & D \\ \hline

\multirow{2}{*}{E/R extractors} & & 
W & W & W &  & & E & E & E & W & & W & W &  & E \\ 
& & 
 &  & D & DM & & M & M & M & M & & M & D & DM & M \\ \hline

Statistical & & 
W & W & W & E & & WE & WE & WE & E & & WE & WE & WE & WE \\ 
components & & 
 &  & D & DM & & M & M & M & DM & & M & M & M & M \\ \hline

Graph  & & 
W & W & W & E & & WE & WE & WE & E & & WE & WE & WE & WE \\ 
components & & 
 &  & D & DM & & M & M & M & DM & & M & M & M & M \\ \hline

\multirow{2}{*}{Query solver} & & 
P & P & W & W & & W & W & W & E & & E & E & E & E \\ 
& & 
 &  &  &  & &  &  & D & DM & & M & M & DM & M \\ \hline

\multirow{2}{*}{Case studies} & & 
W & W & W & W & & W & W & W & W & & W & W & WE & E \\ 
& & 
 &  & D$^1$ &  & &  &  & D$^2$ &  & &  & D$^3$ & D$^4$ M & M \\ \hline

\multirow{2}{*}{Visualization} & & 
P & W & W & W & & E & E & E & E & & E & E & E & E \\ 
& & 
 &  &  & D & & M & M & M & M & & M & M & M & M \\ \hline

\multirow{2}{*}{Integration} & & 
P & W & W & W & & W & W & W & W & & W & W & W & E \\ 
& & 
 &  &  & D & & D & D & D & D & & D & D & D & M \\  \hline

Validation & & 
W & W & W & W & & W & W & W & W & & W & W & W & W \\  \bottomrule
        %\cline{2-13} \\
\end{tabular}
\normalsize
\label{t:workplan}
\end{table}

We will follow an eXtreme Programming 
methodology in the process of building CATARM.
This means that we will define refined incremental items of 
work that allows for more frequent deliverables. 
This also means that the code base will always be subject
to maintenance and refactoring to enhance reuse and 
increase productivity. 
While a item is delivered, it will still be subject to work,
enhancements, and maintenance.

We detailed an agressive work plan in Table~\ref{t:workplan}. 
The plan spans over three years and projects a timeline
of work per quarter. 
The row headers describe work items, and the column headers
describe quarters in the corresponding year of the project.
The legends {\bf P, W, D, M, } 
and {\bf E} 
describe the status of the items 
and stand for 
{\bf P:} {\em planning phase}, 
{\bf W:} {\em work in progress}, 
{\bf D:} {\em deliverable}, 
{\bf M:}  {\em maintainance}, 
and 
{\bf E: } {\em enhance} respectively.

The $1,2,3,$ and $4$
superscripts of the deliverables in the 
in case studies row refer to the four 
case studies described in the propoal.

We will use open source tools, libraries, and frameworks to 
build our libraries.
We will build on existing open source programming and text 
mining tools as well as open source Arabic text analysis tools 
based on NLP such as MADA+TOKAN~\cite{Rot08}, 
NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.

@@

We anticipate that the research will lead to the production of 
CATARM, an Arabic text analysis framework that allows relational 
queries between Arabic text documents.
The components of the framework are listed and discussed
in details in Section~\ref{s:designmethods}.
In the following we will discuss what to anticipate and 
how to evaluate the different components of CATARM.

@@

\subsection{CATARM evaluation }

@@

The completion of the CATARM framework will result in a 
significant contribution to the research community of 
NLP in general and Arabic Analysis in particular. 
The CATARM framework will act as vehicle to enhance and advance 
research in the area. 
We look forward for additional contributions to the field 
such as the following.
\begin{itemize}
\item We will explore developing and integrating vocalization
techniques. 
Vocalization techniques assign diacritics to unvocalized text. 
The prerequisites of a successful vocalizer are the accuracy of 
the morphological analyzer in addition to grammatical analyzers. 
\item We will explore the use of CATARM with more case studies
such as market analysis and opinion mining.
\item We will explore augmenting the lexicon of the
morphological analyzer with useful categorization and entity 
tags that reflect semantical meanings. 
These modifications would give CATARM additional capabilities in 
terms of allowing users to extract and use a greater variety of 
entities when building queries. 
The feedback of experts in the field of Arabic linguistics can 
continually add accuracy to the morphological solutions.  
\item We will explore further optimization techniques
that can help simplify and the analysis. 
\item We will explore generalizing our results to languages 
that resemble the Arabic language. 
Some semetic languages, such as the Amharic, cyrillic, and Hebrew,
share common features with the Arabic 
language. Some other languages are written using the 
Arabic script like Persian, Kurdish, Urdu, and Pashto. 
\item We will also investigate the case-based approach with
Latin languages and check how that compares with current techniques
in terms of accuracy and efficiency.
\item We will maintain CATARM and provide support to users and 
researchers of CATARM.
\end{itemize}
